---
title: "Understanding usdeaths Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding usdeaths Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, eval=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(usdeaths)
library(dplyr)
```

## What Is the Metadata?

The CDC Vital Statistics files are distributed as fixed-width text files —
each row is a long string of characters with no column headers. The only way
to know what those characters mean is to consult the official CDC user guide.

The `usdeaths` package ships that user guide knowledge as structured **metadata
objects** — one per dataset and year. For example, `data_mortality_multiple_2024`
describes the layout of the 2024 multiple-cause-of-death file. These metadata
objects do two things:

1. Tell `read_section()` exactly where to cut each field out of the raw file.
2. Tell `decode_all()` and `decode_preview()` how to translate the CDC's
   numeric codes into human-readable labels.

You rarely need to work with the metadata directly. But understanding its
structure helps you inspect fields, look up what a code means, and troubleshoot
unexpected values in your data.

---

## Metadata Structure

Every metadata object is a **tribble** with one row per field in the raw CDC
file. All metadata objects share the same six columns:

| Column        | Type  | What it tells you                                           |
|---------------|-------|-------------------------------------------------------------|
| `name`        | `chr` | The column name in the parsed tribble                        |
| `start`       | `int` | Starting character position in the raw fixed-width file     |
| `end`         | `int` | Ending character position                                   |
| `size`        | `int` | Number of characters (`end - start + 1`)                    |
| `type`        | `chr` | Storage type — `"int"` or `"str"`                           |
| `description` | `chr` | Plain-English label for the field                           |
| `codes`       | `chr` | Pipe-delimited `value=label` pairs, or `""` if not coded    |

The `codes` column is the heart of the metadata. It stores the CDC's official
value lookups in a compact pipe-delimited format:

```
"1=Residents|2=Intrastate Nonresidents|3=Interstate Nonresidents|4=Foreign Residents"
```

Each entry is a `value=label` pair separated by `|`. A blank string (`""`)
means the field is either free-form text, a reserved filler position, or a
plain numeric quantity that needs no decoding.

---

## Available Metadata Objects

The package includes metadata objects for **mortality multiple cause files**
going back to 1968, as well as birth cohort and births files. The naming
convention is:

```
data_mortality_multiple_[YEAR]   # e.g. data_mortality_multiple_2024
data_birth_cohort_[YEAR]         # e.g. data_birth_cohort_2010
data_births_[YEAR]               # e.g. data_births_2022
```

To see the full list of available years use `get_available_years()`, which returns a tibble of dataset types and years:

```{r available-years, eval=FALSE}
get_available_years("mortality_multiple")
```

---

## Exploring the Metadata

### Exploring Metadata with get_fields()
The easiest way to explore what data is available for a specific year is the get_fields() function. It cleans up the raw metadata by removing "filler" columns and providing a high-level summary.

Listing all available fields
To see what variables are available for the 2024 mortality data:
```{r get-fields, eval=FALSE}
get_fields(type = "mortality", years = 2024) |
```

Searching for specific topics
The Births dataset is much larger (over 200 fields). You can use `get_fields()` to quickly find variables related to a specific topic, like "Tobacco" or "Weight":

```{r, eval=FALSE}
get_fields(type = "births", years = 2024) |> 
  filter(grepl("weight", description, ignore.case = TRUE))
```

Searching over multiple years
One of the most powerful features of `get_fields()` is passing a range of years. This allows you to identify "stable" variables versus those that were recently added or discontinued by the CDC.

```{r, eval=FALSE}
# Compare fields across a 5-year span
fields_comparison <- get_fields(type = "births", years = 2020:2024)

# If 'note' is empty, the field exists in ALL requested years.
# If 'note' contains a year range, the field is only available in those years.
fields_comparison |> 
  filter(note != "") |> 
  select(name, description, note) |> 
  head()
```

### Viewing field names and descriptions

```{r descriptions, eval=FALSE}
data_mortality_multiple_2024 |>
  select(name, start, end, description) |>
  head(10)
```

### Finding fields that have code lookups

Fields where `codes != ""` are the categorical fields that `decode_all()` and
`decode_preview()` will translate for you:

```{r coded-fields, eval=FALSE}
data_mortality_multiple_2024 |>
  filter(codes != "") |>
  select(name, description)
```

### Looking up the codes for a specific field

Use `filter()` and `pull()` to extract the codes string for any field, then
pass it to `decode_column()` to see what values it recognises:

```{r lookup-education, eval=FALSE}
# What codes does the education field use?
edu_codes <- data_mortality_multiple_2024 |>
  filter(name == "education") |>
  pull(codes)

edu_codes
```

You can pass that string directly to `decode_column()` to translate a vector
of raw values:

```{r decode-column-example, eval=FALSE}
decode_column(c(1, 3, 6, 9), edu_codes)
```

`decode_column()` handles leading-zero padding automatically, so integer and
character inputs both work.

---

## From Raw File to Decoded Tribble

The metadata objects sit at the center of every data-loading workflow in the
package. Here is the full picture:

### Option A — One-line import (recommended)

`cdc_import()` is a full pipeline wrapper: it downloads the CDC file,
reads it using the correct metadata, and decodes all coded columns in one call.

```{r cdc-import, eval = FALSE}
# Download, parse, and decode the 2024 mortality data
deaths_2024 <- cdc_import("mortality_multiple", 2024)
```

### Option B — Step by step

If you already have the file on disk, use `read_section()` to parse it and
then `decode_all()` to decode it:

```{r step-by-step, eval = FALSE}
meta <- data_mortality_multiple_2024

# 1. Parse the fixed-width file using the metadata layout
deaths_raw <- read_section("VS24MORT.DUSMCPUB", meta)

# 2. Decode all coded columns in one pass
deaths_decoded <- decode_all(deaths_raw, meta)
```

`read_section()` reads every column as character to avoid silent coercion of
coded numeric fields. `decode_all()` then replaces the raw codes with
human-readable labels and returns columns in metadata order.

### Option C — Preview before committing

While exploring, `decode_preview()` lets you see the original and decoded
values side by side without modifying the full dataset:

```{r decode-preview, eval = FALSE}
# Preview the first 5 coded columns (1 000 rows by default)
decode_preview(deaths_raw, meta)

# Preview 10 coded columns
decode_preview(deaths_raw, meta, first_n = 10)

# Also include column 17 specifically
decode_preview(deaths_raw, meta, numbers = 17)

# Decode everything and return the full dataset
deaths_decoded <- decode_preview(deaths_raw, meta, all = TRUE, n_rows = nrow(deaths_raw))
```

When `all = FALSE` (the default), `decode_preview()` returns each original
column immediately next to a `_label` companion column — convenient for
spot-checking that codes are mapping as expected before running `decode_all()`.

---

## Field Groups at a Glance

The metadata covers several thematic groups. Here is a quick overview of the
most commonly used ones.

### Administrative / record identifiers

`record_type` and `resident_status` identify whether the decedent was a state
resident or came from out of state. These are important when filtering to a
specific geographic population.

```{r admin-fields, eval=FALSE}
data_mortality_multiple_2024 |>
  filter(name %in% c("record_type", "resident_status")) |>
  select(name, description, codes)
```

### Education

Two fields capture education: the level itself and a reporting flag indicating
which version of the death-certificate education question was used. Because
states adopted the 2003 revision at different times, the flag is essential for
longitudinal comparisons.

```{r education-fields, eval=FALSE}
data_mortality_multiple_2024 |>
  filter(startsWith(name, "education")) |>
  select(name, description, codes)
```

### Reserved / filler fields

Fields named `reserved_*` are placeholder positions in the fixed-width layout.
They carry no information and can be safely ignored in analysis, but the
metadata includes them so that all character positions remain correct for
`read_section()`.

```{r reserved-fields, eval=FALSE}
data_mortality_multiple_2024 |>
  filter(startsWith(name, "reserved")) |>
  select(name, start, end, size)
```

---

## Tips for Beginners

- **`start` and `end` positions** are 1-indexed character positions in the
  raw flat file. You don't normally need them once the file has been parsed,
  but they're useful if you're debugging a parsing issue.

- **`type = "int"` vs `"str"`** indicates how the raw characters should be
  interpreted. The CDC conventionally uses `"9"`, `"99"`, `"999"`, etc. to
  represent "Unknown" for fields of different widths — always check the
  `codes` column before treating those values as numeric unknowns.

- **`decode_column()` for one-off lookups** — if you just need to translate a
  handful of values interactively, `decode_column()` takes any vector and a
  codes string and returns labels directly, without needing a full dataset.

- **Match metadata year to data year** — always use the metadata object that
  corresponds to your data year. The CDC's field layouts and code definitions
  change across years, so `data_mortality_multiple_2010` and
  `data_mortality_multiple_2024` are not interchangeable.

---